Sperm do not evolve to collaborate in female meiotic drive.

### 2014-05-07 writeup_and_notes/male_dependent_drive.tex
updated text on bistability, fixed bug in additive bistability

The positive frequency dependence of a self-promoting driver can
induce bistability in its spread, i.e. there is a critical frequency
above which it can spread and below which it moves towards loss. 
Compared to the invasion criteria, the bistable parameter space allowing fixations is much broader -- 
	a common, recessive, self-promoting driver fixes if
        $s<(2d-1)/2$ approximately. 
Therefore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, 
	a recessive, self-promoting driver will deterministically fix if drift, mutation, 
	or migration pressure bring its frequency above $(1-2d+4ds)/(2s(2d-1))$, 
	but it will be lost otherwise.
Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-h_s$)
	further constrains the evolution of a self-promoting driver. 
This case also displays bistability -- with any heterozygous fitness cost, a rare self-promoting driver cannot spread. 
However, if $s < d - 1/2+s_h(3-2d)/2$, 
	it will fix deterministically if its 
	frequency is greater than a threshold, $f_B^*$ [Equation \eqref{thresh2}, and \yb{FIGURES SXX}].
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive})
%While a standard female driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
%	a rare self-promoter pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
%This frequency-independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Below a cutoff frequency the self-promoter driver deterministically decreases in frequency 
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive}). 
%Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
%	make drive effective enough to overcome the associated fitness cost.
This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that, if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be
observed as stable polymorphisms in natural populations.

@@

\yb{\paragraph{Recessive fitness cost of self-promoting driver}
When the self-promoting driver has a recessive fitness cost, $s$ genotype frequencies after selection and drive equal
\begin{equation}	x'_{BB} =\frac{x_{B}^2(1-s)}{\bar{w} }  \Big( F+(1-F)(x_{B}+2x_{A} d) \Big) 	\end{equation}
\begin{equation}	x'_{AA} =\frac{1}{\bar{w} }  \Big(x_{A}^2\Big) 	\end{equation}
and
\begin{equation}	x'_{AB} =\frac{2 x_{A}x_{B} }{\bar{w} }  \Big(1-x_{B}(d-1/2)(1-F) \Big)	\end{equation}

its change in frequency across generations $\Delta f_B$ is
\begin{equation}
\frac{f_A f_B^2}{\bar{w}} \left( [1-F][ d(1-2 f_A s) - f_B s-1/2] - s  F \right)
\label{deltadriver}
\end{equation}
where $F$ is the deviation from Hardy-Weinberg expectations, and $f_A$ is the frequency of the non-driving allele. 
For analytical solutions presented above, we assume that $F$ equals zero. 
}

\paragraph{cost of driver in heterozygotes}
We explored a more complex model where the self-promoting driver has a cost, $s_h$ in heterozygotes (so $w_{AB} = 1-h_s$) in addition to the cost $s$, in homozygotes.
Model construction is similar to that above.
So long as there is any heterozygous fitness cost (i.e. $s_h > 0$) the self-promoting driver cannot invade when rare;
	however, as is the case without a heterozygous cost, this system is bistable.
Assuming no deviation from Hardy-Weinberg a self-promoting driver with a heterozygous fitness cost deterministically fixes if it's frequency exceeds some threshold, $f_B^*$.

\begin{equation}
	f_B^* = \frac{a_1+\sqrt{a_1^2+8 s_h a_2}}{2 a_2}
	\label{thresh2}
\end{equation}
where, $a_1=1-2 ds_h+4 d s-2 d-3 s_h$ and $a_2=2(s-s_h)(2d-1)$.

@@

\paragraph{Dominance} In the supplement we also explore a model where
the genotype of the male who produced the fertilizing sperm affects female drive,
instead of just the allelic type of the sperm.
 We assume a heterozygous female will transmit the $B$ allele 
  with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $ or $d$, 
 if she mated with a n $AA$, $AB$, or $BB$ male,  respectively.
 Above, $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
%YB TRY_1 We assume that in a mating between a heterozygous female and an $AA$, $AB$, or $BB$ male, the female will transmit the $B$ allele 
% with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $, and $d$, respectively.
  %Here $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
% GC ORIGINAL  We assume that if a
%heterozygous egg ($AB$) is fertilized by a sperm with the following
%male ($\Mars$) genotypes the $B$ allele will be transmitted through
%female meiosis with the following probabilities \\
%\begin{tabular}{cccc}
%$\Mars$ geno. & AA & AB  & BB\\
%P(B). & $\frac{1}{2}$ &  $\frac{1}{2} + h(d-\frac{1}{2}) $ & $d$ 
%\end{tabular}\\
%where $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype. 
We set up a set of recursions using a recessive sex independent cost to
the driver, similar to those in eqn.
\ref{eqn.A1}-\ref{eqn.A1}, but now featuring these male genotype
transmission probabilites. In Supplementary Figure \ref{Effect_of_dominance} we show the invasion
criteria for both the A and B alleles when they are rare. 
We find a narrow parameter range where B 
can invade but not fix.
This corresponds to B being less than fully recessive (i.e. $h>0$) and to a modest cost in homozygotes. 
This combination of parameters reflects the fact that when the allele is rare, 
it can increase in frequency due to fertilization of heterozygous eggs
by sperm from heterozygote males, where the sperm do not themselves
carry the allele. But as the B allele becomes common it will frequently be
present in the homozygous state in males, and so will be unable to
avoid the production of unfit homozygote zygotes leaving it trapped in
the population. \gc{Perhaps we should show the equil. freq. in the
  fig, as I'm guessing it might be rare}

@@

\gc{Alternatively we could give the table below:}

\begin{table}
\begin{tabular}{lllll}
\hline 
  &  \multicolumn{4}{c}{Sperm genotype}   \\
  &  & A & B & B'  \\
$\Venus$ geno. & Freq.  & $x_{AA} + \frac{1}{2} x_{AB}+ \frac{1}{2} x_{AB'} $ & $ w_{BB}  x_{BB} +   w_{BB'} \frac{1}{2} x_{BB'}+
\frac{1}{2} x_{AB}$ & $ w_{B'B'}  x_{B'B'} +   w_{BB'} \frac{1}{2}
x_{BB'}+ \frac{1}{2} x_{AB'}  $\\
\hline\\
AA & $x_{AA}$ &  & & \\
AB & $x_{AB}$ &  & & \\
AB' & $x_{AB'}$ &  & & \\
BB' & $w_{BB'} x_{BB'}$ &  & & \\
BB & $w_{B'B'} x_{BB}$ &  & & \\
B'B' & $w_{B'B'} x_{B'B'}$ &  & & \\
\end{tabular}
\caption{ALTERNATIVELY I COULD FILL IN THIS MATRIX  with genotype }
\end{table}
\bibliography{refs}

@@

@@

\section*{OUTLINE}
\yb{YB: To me its (1) male enhancement of female drive cannot maintain a stable polymorphisms, and (2) An allele in sperm can evolve to suppress its drive in females.}
show that such alleles:
\begin{enumerate}
\item can't be balanced, \\
\item and homozygous problems are tested out at low freq.  \\
\item Any heterozygous problems, leads to a bistable allele\\
\item If these alleles take off they speed through to fixation\\
\item If allele has any drive ability in absence of sperm effect that is what allows it to enter the population
and sperm effect isn't a further cause of conflict. What if anything do we mean by this?\\
\item PERHAPS HERE WE INTRODUCE A SELF-RESTRAINING ALLELE
\end{enumerate}

Conclusion, such alleles are unlikely to cause evolution of female supressors, they test hemselves in a homozygous
state when they enter the population, and sweep quickly (all the way to fixation) if they enter the pop at all.\\

\begin{enumerate}
\item Setup a drive-selection balanced polymorphism in std. drive model. 
Do this by imagining the sperm-influence allele arising on the background of the driver, 
so the allele has drive capabilities, and can have had time to evolve new biology.

Multilocus drivers held together with inversions, may offer a relatively large mutational/functional target.

Evolving on the new background means that the allele suffers the fitness consequences of the 
driver.  See A in Figure \ref{Eggsperm_3_allele_cartoon} \\
\item Sperm-based enhancers of drive can't invade (can they in some situations?). \\
\item Intuition is that the driver has already driven to a frequency 
where it is held in check by its cost in homozygotes. The sperm allele 
thus can't really help as it creates zygotes which suffer the homozygous fitness consequences.
\end{enumerate}

\subsection{So what can evolve?}
So what can happen?
\begin{enumerate}
\item Alleles that arise in linkage with drive systems, which when in sperm switch off drive, 
can spread. They benefit from drive, but avoid some of the
consequences ( (Bi) in Figure \ref{Eggsperm_3_allele_cartoon} ). Overall as a side product they are benefiting all in pop.\\
\item Presumably alleles that actually switch the allele that drives may do even better? As they'd end up in 
hets. Although they'd not drive, so hard to say. YB: There is evidence that In distorts meiosis in the other direction (they still drive when rare i.e. when not fused with drive supp sperm) \\
\item Alleles that cause sperm to switch off drive that arise on other background or unlinked to the drive system
are selected, and spread as fast as female supressors of drive ( (Bii) in Figure \ref{Eggsperm_3_allele_cartoon} )..\\
\item Alleles that in females facilitate the action of sperm supressors of drive (or vis versa) can spread. Haven't actually checked this.\\
\end{enumerate}

\section*{Conclusions.}

>>>>

The positive frequency dependence of a self-promoting driver can
induce bistability in its spread, i.e. there is a critical frequency
above which it can spread quickly to fixation and below which it moves towards loss (an
unstable equilibrium frequency). In 
the appendix we devlop a simple
approximation to this bistability 
This bistability occurs in the region of parameter space in Figure
\ref{Invasion_space}. In Supplementary Figure
\ref{Bistab_homozyg_cost_fig} we further explore this bistability and
show that the unstable equilibrium frequency is high across much of
this parameter space, and so in practice the self promoting allele
only invade in the orange space. (In the appendix we develop a simple
approximation to the unstable equilibrium frequency
\eqref{bistab_homozy_approx} that is also shown in  Supplementary Figure
\ref{Bistab_homozyg_cost_fig}).
% Compared to the invasion criteria, the parameter space
% allowing the fixation of the allele if it is above a cutoff frequency is much broader -- 
% 	a common, recessive, self-promoting driver fixes if
%         $s<(2d-1)/2$ approximately. 
% Therefore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, 
% 	a recessive, self-promoting driver will deterministically fix if drift, mutation, 
% 	or migration pressure bring its frequency above $(1-2d+4ds)/(2s(2d-1))$, 
% 	but it will be lost otherwise.
Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-h_s$)
strongly constrains the evolution of a self-promoting driver. 
In fact with any heterozygous fitness cost, a rare self-promoting
driver cannot spread deterministically. However, this case also displays bistability -- 
if $s < d - 1/2+s_h(3-2d)/2$ approximately, it will fix deterministically if its 
	frequency is greater than a threshold, $f_B^*$ [Equation \eqref{thresh2}, and \yb{FIGURES SXX}].
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive})
%While a standard female driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
%	a rare self-promoter pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
%This frequency-independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Below a cutoff frequency the self-promoter driver deterministically decreases in frequency 
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive}). 
%Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
%	make drive effective enough to overcome the associated fitness cost.
This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that, if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be
observed as stable polymorphisms in natural populations.

@@

\paragraph{Recessive fitness cost of self-promoting driver}
When the self-promoting driver has a recessive fitness cost, $s$ genotype frequencies after selection and drive equal
\begin{equation}	x'_{BB} =\frac{x_{B}^2(1-s)}{\bar{w} }  \Big( F+(1-F)(x_{B}+2x_{A} d) \Big) 	\end{equation}
\begin{equation}	x'_{AA} =\frac{1}{\bar{w} }  \Big(x_{A}^2\Big) 	\end{equation}
and
\begin{equation}	x'_{AB} =\frac{2 x_{A}x_{B} }{\bar{w} }  \Big(1-x_{B}(d-1/2)(1-F) \Big)	\end{equation}

its change in frequency across generations $\Delta f_B$ is
\begin{equation}
\frac{f_A f_B^2}{\bar{w}} \left( [1-F][ d(1-2 f_A s) - f_B s-1/2] - s  F \right)
\label{deltadriver}
\end{equation}
where $F$ is the deviation from Hardy-Weinberg expectations (HWE), and $f_A$ is the frequency of the non-driving allele. 
For analytical solutions presented above, we assume that $F$ equals zero.

We used this HWE approximation to investifgate the bistability of the
invasion criteria. Under this approximation we find that a common, recessive, self-promoting driver fixes if
 $s<(2d-1)/2$. Furthermore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, a
 recessive, self-promoting driver will deterministically fix if drift,
 mutation, or migration pressure bring its frequency above approximately
\begin{equation} 
(1-2d+4ds)/(2s(2d-1)) \label{bistab_homozy_approx}
\end{equation}
 but it will be lost if it is introduced below this frequency
 (i.e. this is an unstable equilibrium point). In Supplementary Figure \ref{Bistab_homozyg_cost_fig} we show
 this approximation condition for bistability andd how it compared to
 that obtained using the recursion. This equation
 \eqref{bistab_homozy_approx} offers a reasonable, but imperfect
 approximation 
and seems to serve as a lower bound to the
true unstable equilibrium frequency.

\paragraph{cost of driver in heterozygotes}
We explored a more complex model where the self-promoting driver has a cost, $s_h$ in heterozygotes (so $w_{AB} = 1-h_s$) in addition to the cost $s$, in homozygotes.
Model construction is similar to that above.
So long as there is any heterozygous fitness cost (i.e. $s_h > 0$) the self-promoting driver cannot invade when rare;
	however, as is the case without a heterozygous cost, this system is bistable.
Assuming no deviation from Hardy-Weinberg a self-promoting driver with a heterozygous fitness cost deterministically fixes if it's frequency exceeds some threshold, $f_B^*$.
\begin{equation}
	f_B^* = \frac{a_1+\sqrt{a_1^2+8 s_h a_2}}{2 a_2}
	\label{thresh2}
\end{equation}
where, $a_1=1-2 ds_h+4 d s-2 d-3 s_h$ and $a_2=2(s-s_h)(2d-1)$.

@@

\paragraph{Dominance} we also explored a model where
the genotype of the male who produced the fertilizing sperm affects female drive,
instead of just the allelic type of the sperm.
 We assume a heterozygous female will transmit the $B$ allele 
  with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $ or $d$, 
 if she mated with a n $AA$, $AB$, or $BB$ male,  respectively.
 Above, $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
%YB TRY_1 We assume that in a mating between a heterozygous female and an $AA$, $AB$, or $BB$ male, the female will transmit the $B$ allele 
% with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $, and $d$, respectively.
  %Here $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
% GC ORIGINAL  We assume that if a
%heterozygous egg ($AB$) is fertilized by a sperm with the following
%male ($\Mars$) genotypes the $B$ allele will be transmitted through
%female meiosis with the following probabilities \\
%\begin{tabular}{cccc}
%$\Mars$ geno. & AA & AB  & BB\\
%P(B). & $\frac{1}{2}$ &  $\frac{1}{2} + h(d-\frac{1}{2}) $ & $d$ 
%\end{tabular}\\
%where $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype. 
We set up a set of recursions using a recessive sex independent cost to
the driver, similar to those in eqn.
\ref{eqn.A1}-\ref{eqn.A1}, but now featuring these male genotype
transmission probabilites. In Supplementary Figure \ref{Effect_of_dominance} we show the invasion
criteria for both the A and B alleles when they are rare. 
We find a narrow parameter range where B 
can invade but not fix.
This corresponds to B being less than fully recessive (i.e. $h>0$) and to a modest cost in homozygotes. 
This combination of parameters reflects the fact that when the allele is rare, 
it can increase in frequency due to fertilization of heterozygous eggs
by sperm from heterozygote males, where the sperm do not themselves
carry the allele. But as the B allele becomes common it will frequently be
present in the homozygous state in males, and so will be unable to
avoid the production of unfit homozygote zygotes leaving it trapped in
the population. \gc{Perhaps we should show the equil. freq. in the
  fig, as I'm guessing it might be rare}

@@

% \gc{Alternatively we could give the table below:}

% \begin{table}
% \begin{tabular}{lllll}
% \hline 
%   &  \multicolumn{4}{c}{Sperm genotype}   \\
%   &  & A & B & B'  \\
% $\Venus$ geno. & Freq.  & $x_{AA} + \frac{1}{2} x_{AB}+ \frac{1}{2} x_{AB'} $ & $ w_{BB}  x_{BB} +   w_{BB'} \frac{1}{2} x_{BB'}+
% \frac{1}{2} x_{AB}$ & $ w_{B'B'}  x_{B'B'} +   w_{BB'} \frac{1}{2}
% x_{BB'}+ \frac{1}{2} x_{AB'}  $\\
% \hline\\
% AA & $x_{AA}$ &  & & \\
% AB & $x_{AB}$ &  & & \\
% AB' & $x_{AB'}$ &  & & \\
% BB' & $w_{BB'} x_{BB'}$ &  & & \\
% BB & $w_{B'B'} x_{BB}$ &  & & \\
% B'B' & $w_{B'B'} x_{B'B'}$ &  & & \\
% \end{tabular}
% \caption{ALTERNATIVELY I COULD FILL IN THIS MATRIX  with genotype }
% \end{table}
\bibliography{refs}

@@

\begin{figure}
	\rotatebox{90}{
\includegraphics[width = 0.5
          \textwidth]{Scripts/homozyg_bistability.eps}}
\caption{Invasion analysis for a self promoting female meiotic drive allele with
  recessive costs (selection coefficient $s$), showing the region of
  bistability. The colors, and the thin dashed contours, indicate the
  frequency the allele must be above to invade the population (note
  that these alleles reach fixation conditional on invading). In the
  white area the allele can not invade ($x \geq1$), in the solid red
  area the allele can invade and fix when rare. In the left panel we
  show the results obtained by a grid search using the recursion, on
  the right we show the approximation obtained assuming that HWE
  holds. }  
\label{Bistab_homozyg_cost_fig}
\end{figure}

@@
